---
title: "TileDB Dimensions: Date and Datetime Support"
output:
  minidown::mini_document:
    framework: water
---

```{r include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      class.output = NULL,
                      class.message = NULL,
                      class.warning = NULL,
                      class.error = NULL,
                      python.reticulate = TRUE)  # can toggle to off
Sys.setenv("TZ"="UTC")
```


## Overview

TileDB supports a wide range of date and time types as shown in the [documention for
Datetimes][doc_datetimes].

The supported range of dates and times correspond to the dates and times supported by the Python
extension *numpy*, and is described [in its documentation][numpy_datetime].

```{python pyex1}
import numpy as np
import datetime

day = "1971-02-03 04:05:06"    # third day of second month of first year past epoch

[np.datetime64(day, 'Y').astype('int64'),    # extract year, month, ... as an int64
np.datetime64(day, 'M').astype('int64'),
np.datetime64(day, 'D').astype('int64'),
np.datetime64(day, 'h').astype('int64'),
np.datetime64(day, 'm').astype('int64'),
np.datetime64(day, 's').astype('int64')]
```

R has date and datetime support built-in.  The `Date` supports dates using an integer count since
the epoch, and matches the `D` value from the previous example.  Similarly, the 'compact' `POSIXct`
representation of a `Datetime` uses the number of seconds since the epoch and corresponds to the `s`
value from the previous example.


```{r rex1}
as.Date(398, origin="1970-01-01")  # conversion from numeric input requires 'origin'
as.POSIXct(34401906, origin="1970-01-01", tz="UTC", usetz=TRUE)
```


R can reconstruct dates and times from the *numpy* representation using the epoch as a 'base' date
along with time period calculations. This can be done using base R (adding to `Date` or `POSIXct`
objects), by using the *lubridate* package for a number of intermediate formats and by using the
*nanotime* package for higher-resolution periods and intervals.

```{r rex1a}
suppressMessages(library(lubridate))
ymd("1970-01-01") +  c(years(1), months(13), days(398))
ymd_hms("1970-01-01 00:00:00") + c(hours(9556), minutes(573365))

suppressMessages(library(nanotime))
nanotime("1970-01-01T00:00:00+00:00") + nanoduration(hours=2, minutes=3,
                                                     seconds=4, nanoseconds=5)
```

<!-- Note that the is result is correct to the given precision: adding one year to the epoch gets us a
date in 1971 (here the first one). Similarly, adding thirteen months gets us to the correct month
but not yet date. A day can only be fully recovered via the 'days' period, here 398.
-->


## Python and R Interoperability

### Coarsest: Year

#### Python

```{python pyex2}
import numpy as np
import sys
import os
import tiledb
uri = "/tmp/tiledb/dt_year"
dom = tiledb.Domain(tiledb.Dim(name="rows",
                               domain=(np.datetime64('2001-01-01'), np.datetime64('2030-12-31')),
                               tile=np.timedelta64(10, 'Y'),
                               dtype=np.datetime64('', 'Y')))
schema = tiledb.ArraySchema(domain=dom,
                            sparse=True,
                            attrs=[tiledb.Attr(name="a", dtype=np.int32)])
if (os.path.isdir(uri)):
    tiledb.VFS().remove_dir(uri)
tiledb.SparseArray.create(uri, schema)

with tiledb.SparseArray(uri, mode='w') as A:
    I = [np.datetime64('2001-01-01'),np.datetime64('2002-01-01'),np.datetime64('2003-01-01')]
    data = np.array(([1,2,3]))
    A[I] = data


```

#### R

```{r rex2}
library(tiledb)
uri <- "/tmp/tiledb/dt_year"
arr <- tiledb_array(uri, as.data.frame=TRUE)
arr[]
## we can also look at 'raw' int64 values:
datetimes_as_int64(arr) <- TRUE
arr[]
```



### Day

#### Python

```{python pyex3}
import numpy as np
import sys
import os
import tiledb
uri = "/tmp/tiledb/dt_day"
dom = tiledb.Domain(tiledb.Dim(name="rows",
                               domain=(np.datetime64('2001-01-01'), np.datetime64('2030-12-31')),
                               tile=np.timedelta64(10, 'D'),
                               dtype=np.datetime64('', 'D')))
schema = tiledb.ArraySchema(domain=dom,
                            sparse=True,
                            attrs=[tiledb.Attr(name="a", dtype=np.int32)])
if (os.path.isdir(uri)):
    tiledb.VFS().remove_dir(uri)
tiledb.SparseArray.create(uri, schema)

with tiledb.SparseArray(uri, mode='w') as A:
    I = [np.datetime64('2001-01-01'),np.datetime64('2001-01-02'),np.datetime64('2001-01-03')]
    data = np.array(([1,2,3]))
    A[I] = data

```

#### R

```{r rex3}
library(tiledb)
uri <- "/tmp/tiledb/dt_day"
arr <- tiledb_array(uri, as.data.frame=TRUE)
arr[]
```



### Minute

#### Python

```{python pyex4}
import numpy as np
import sys
import os
import tiledb
uri = "/tmp/tiledb/dt_min"
dom = tiledb.Domain(tiledb.Dim(name="rows",
                               domain=(np.datetime64('2001-01-01'), np.datetime64('2030-12-31')),
                               tile=np.timedelta64(10, 'm'),
                               dtype=np.datetime64('', 'm')))
schema = tiledb.ArraySchema(domain=dom,
                            sparse=True,
                            attrs=[tiledb.Attr(name="a", dtype=np.int32)])
if (os.path.isdir(uri)):
    tiledb.VFS().remove_dir(uri)
tiledb.SparseArray.create(uri, schema)

with tiledb.SparseArray(uri, mode='w') as A:
    I = [np.datetime64('2001-01-01 00:00'),
         np.datetime64('2001-01-02 00:01'),
         np.datetime64('2001-01-03 00:02')]
    data = np.array(([1,2,3]))
    A[I] = data

```

#### R

```{r rex4}
library(tiledb)
uri <- "/tmp/tiledb/dt_min"
arr <- tiledb_array(uri, as.data.frame=TRUE)
arr[]
```





### Millisecond

#### Python

```{python pyex5a}
import numpy as np
import sys
import os
import tiledb
uri = "/tmp/tiledb/dt_ms"
dom = tiledb.Domain(tiledb.Dim(name="rows",
                               domain=(np.datetime64('1969-01-01'), np.datetime64('2030-12-31')),
                               tile=np.timedelta64(10, 'ms'),
                               dtype=np.datetime64('', 'ms')))
schema = tiledb.ArraySchema(domain=dom,
                            sparse=True,
                            attrs=[tiledb.Attr(name="a", dtype=np.int32)])
if (os.path.isdir(uri)):
    tiledb.VFS().remove_dir(uri)
tiledb.SparseArray.create(uri, schema)

with tiledb.SparseArray(uri, mode='w') as A:
    I = [np.datetime64('1970-01-01 00:00:00.001'),
         np.datetime64('1980-01-01 00:00:00.002'),
         np.datetime64('1990-01-01 00:00:00.003'),
         np.datetime64('2000-01-01 00:00:00.004'),
         np.datetime64('2010-01-01 00:00:00.005'),
         np.datetime64('2020-01-01 00:00:00.006')]
    data = np.array(([1,2,3,4,5,6]))
    A[I] = data

```

#### R

```{r rex5a}
library(tiledb)
uri <- "/tmp/tiledb/dt_ms"
arr <- tiledb_array(uri, as.data.frame=TRUE)
arr[]
```



### Microsecond

#### Python

```{python pyex5}
import numpy as np
import sys
import os
import tiledb
uri = "/tmp/tiledb/dt_us"
dom = tiledb.Domain(tiledb.Dim(name="rows",
                               domain=(np.datetime64('1969-01-01'), np.datetime64('2030-12-31')),
                               tile=np.timedelta64(10, 'us'),
                               dtype=np.datetime64('', 'us')))
schema = tiledb.ArraySchema(domain=dom,
                            sparse=True,
                            attrs=[tiledb.Attr(name="a", dtype=np.int32)])
if (os.path.isdir(uri)):
    tiledb.VFS().remove_dir(uri)
tiledb.SparseArray.create(uri, schema)

with tiledb.SparseArray(uri, mode='w') as A:
    I = [np.datetime64('1970-01-01 00:00:00.000001'),
         np.datetime64('1980-01-01 00:00:00.000002'),
         np.datetime64('1990-01-01 00:00:00.000003'),
         np.datetime64('2000-01-01 00:00:00.000004'),
         np.datetime64('2010-01-01 00:00:00.000005'),
         np.datetime64('2020-01-01 00:00:00.000006')]
    data = np.array(([1,2,3,4,5,6]))
    A[I] = data

```

#### R

```{r rex5}
library(tiledb)
uri <- "/tmp/tiledb/dt_us"
arr <- tiledb_array(uri, as.data.frame=TRUE)
arr[]
```




### Nanosecond

#### Python

```{python pyex6}
import numpy as np
import sys
import os
import tiledb
uri = "/tmp/tiledb/dt_ns"
dom = tiledb.Domain(tiledb.Dim(name="rows",
                               domain=(np.datetime64('1969-01-01'), np.datetime64('2030-12-31')),
                               tile=np.timedelta64(10, 'ns'),
                               dtype=np.datetime64('', 'ns')))
schema = tiledb.ArraySchema(domain=dom,
                            sparse=True,
                            attrs=[tiledb.Attr(name="a", dtype=np.int32)])
if (os.path.isdir(uri)):
    tiledb.VFS().remove_dir(uri)
tiledb.SparseArray.create(uri, schema)

with tiledb.SparseArray(uri, mode='w') as A:
    I = [np.datetime64('1970-01-01 00:00:00.000000001'),
         np.datetime64('1980-01-01 00:00:00.000000002'),
         np.datetime64('1990-01-01 00:00:00.000000003'),
         np.datetime64('2000-01-01 00:00:00.000000004'),
         np.datetime64('2010-01-01 00:00:00.000000005'),
         np.datetime64('2020-01-01 00:00:00.000000006')]
    data = np.array(([1,2,3,4,5,6]))
    A[I] = data

```

#### R

```{r rex6}
library(tiledb)
uri <- "/tmp/tiledb/dt_ns"
arr <- tiledb_array(uri, as.data.frame=TRUE)
arr[]
```


## Use integer64 Directly

Sometimes we may want to access the date or datetimes value in their native
`integer64` format. To do so, we set a toggle when opening the array as shown in the following
example which uses the array from the preceding example (at resolution of nanosecond).

```{r rex7}
library(tiledb)
uri <- "/tmp/tiledb/dt_ns"
arr <- tiledb_array(uri, as.data.frame=TRUE, datetimes_as_int64=TRUE)
arr[]
```

We can also write `integer64` types. The following example adds two extra rows:

```{r rex8}
library(tiledb)
uri <- "/tmp/tiledb/dt_ns"
arr <- tiledb_array(uri, as.data.frame=TRUE, datetimes_as_int64=TRUE)
arr[] <- data.frame( rows=bit64::as.integer64(2:3), a=102:103)
arr[]
```



[doc_datetimes]: https://docs.tiledb.com/main/solutions/tiledb-embedded/internal-mechanics/datetimes
[numpy_datetime]: https://numpy.org/doc/stable/reference/arrays.datetime.html
